---
title: "Diabetes Risk Prediction Using Behavioral and Demographic Factors"
author: "daisy_gui"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cerulean
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE, 
  warning = FALSE
)
```

# Motivation

Diabetes is one of the most prevalent chronic diseases in the United States, affecting over 34 million Americans and placing an estimated $327 billion annual burden on the healthcare system. 

The disease reduces life expectancy and quality of life while contributing to severe complications including heart disease, kidney failure, vision loss, and amputations. 

A significant challenge in diabetes management is underdiagnosis. 1 in 5 diabetics and 8 in 10 prediabetics are unaware of their condition. Therefore, **early identification of at-risk individuals** enables lifestyle modification and preventive care, potentially mitigating the burden of diabetes. 

```{r}
library(tidyverse)
library(caret)
library(skimr)
library(janitor)
library(corrplot)
library(DataExplorer)
library(ggplot2)
library(pROC)
library(broom)
set.seed(123)
```

```{r}
diabetes_raw <- read_csv("data/diabetes_binary_health_indicators_BRFSS2015.csv")
```

```{r}
diabetes_clean <- diabetes_raw %>%
clean_names() %>%
rename(heart_disease_or_attack = heart_diseaseor_attack) %>%
distinct() %>%
drop_na()
```

```{r}
diabetes_processed <- diabetes_clean %>%
mutate(
diabetes_binary = factor(diabetes_binary,
levels = c(0, 1),
labels = c("No_Diabetes", "Diabetes")),

high_bp = factor(high_bp, levels = c(0, 1), labels = c("No", "Yes")),
high_chol = factor(high_chol, levels = c(0, 1), labels = c("No", "Yes")),
chol_check = factor(chol_check, levels = c(0, 1), labels = c("No", "Yes")),
smoker = factor(smoker, levels = c(0, 1), labels = c("No", "Yes")),
stroke = factor(stroke, levels = c(0, 1), labels = c("No", "Yes")),
heart_disease_or_attack = factor(heart_disease_or_attack, 
                                 levels = c(0, 1), 
                                 labels = c("No", "Yes")),
phys_activity = factor(phys_activity, levels = c(0, 1), labels = c("No", "Yes")),
fruits = factor(fruits, levels = c(0, 1), labels = c("No", "Yes")),
veggies = factor(veggies, levels = c(0, 1), labels = c("No", "Yes")),
hvy_alcohol_consump = factor(hvy_alcohol_consump, levels = c(0, 1), labels = c("No", "Yes")),
any_healthcare = factor(any_healthcare, levels = c(0, 1), labels = c("No", "Yes")),
no_docbc_cost = factor(no_docbc_cost, levels = c(0, 1), labels = c("No", "Yes")),
diff_walk = factor(diff_walk, levels = c(0, 1), labels = c("No", "Yes")),
sex = factor(sex, levels = c(0, 1), labels = c("Female", "Male")),

gen_hlth_cat = factor(gen_hlth,
                      levels = 1:5,
                      labels = c("Excellent", "Very Good", "Good", "Fair", "Poor"),
                      ordered = TRUE),

age_cat = factor(age,
                 levels = 1:13,
                 labels = c("18-24", "25-29", "30-34", "35-39", "40-44",
                            "45-49", "50-54", "55-59", "60-64", "65-69",
                            "70-74", "75-79", "80+"),
                 ordered = TRUE),

education_cat = factor(education,
                       levels = 1:6,
                       labels = c("Never/K-8", "Grades 9-11", "Grade 12/GED",
                                  "College 1-3 yrs", "College 4+ yrs", "Unknown"),
                       ordered = TRUE),

income_cat = factor(income,
                    levels = 1:8,
                    labels = c("<$10k", "$10k-$15k", "$15k-$20k", "$20k-$25k",
                               "$25k-$35k", "$35k-$50k", "$50k-$75k", "$75k+"),
                    ordered = TRUE),

bmi_category = case_when(
  bmi < 18.5 ~ "Underweight",
  bmi >= 18.5 & bmi < 25 ~ "Normal",
  bmi >= 25 & bmi < 30 ~ "Overweight",
  bmi >= 30 ~ "Obese"
) %>% factor(levels = c("Underweight", "Normal", "Overweight", "Obese"),
             ordered = TRUE),

ment_hlth_cat = case_when(
  ment_hlth == 0 ~ "None",
  ment_hlth <= 7 ~ "1-7 days",
  ment_hlth <= 14 ~ "8-14 days",
  ment_hlth <= 30 ~ "15-30 days"
) %>% factor(levels = c("None", "1-7 days", "8-14 days", "15-30 days"),
             ordered = TRUE),

phys_hlth_cat = case_when(
  phys_hlth == 0 ~ "None",
  phys_hlth <= 7 ~ "1-7 days",
  phys_hlth <= 14 ~ "8-14 days",
  phys_hlth <= 30 ~ "15-30 days"
) %>% factor(levels = c("None", "1-7 days", "8-14 days", "15-30 days"),
             ordered = TRUE),

health_risk_score = as.numeric(high_bp == "Yes") +
  as.numeric(high_chol == "Yes") +
  as.numeric(stroke == "Yes") +
  as.numeric(heart_disease_or_attack == "Yes") +
  as.numeric(smoker == "Yes") +
  as.numeric(hvy_alcohol_consump == "Yes") +
  (gen_hlth - 1) / 4,

lifestyle_score = as.numeric(phys_activity == "Yes") +
  as.numeric(fruits == "Yes") +
  as.numeric(veggies == "Yes")


)
```

```{r}
# Train/test split
set.seed(123)

train_index <- createDataPartition(
diabetes_processed$diabetes_binary,
p = 0.80,
list = FALSE
)

diabetes_train <- diabetes_processed[train_index, ]
diabetes_test  <- diabetes_processed[-train_index, ]
```

```{r}
# Export Processed Datasets
if (!dir.exists("data")) dir.create("data")

write_csv(diabetes_processed, "data/diabetes_processed.csv")
write_csv(diabetes_train, "data/diabetes_train.csv")
write_csv(diabetes_test,  "data/diabetes_test.csv")

saveRDS(diabetes_processed, "data/diabetes_processed.rds")
saveRDS(diabetes_train, "data/diabetes_train.rds")
saveRDS(diabetes_test,  "data/diabetes_test.rds")
```

```{r}
# data directionary
data_dictionary <- tribble(
~Variable, ~Type, ~Description, ~Values,
"diabetes_binary", "Factor", "Diabetes diagnosis", "No_Diabetes, Diabetes",
"high_bp", "Factor", "High blood pressure", "No, Yes",
"high_chol", "Factor", "High cholesterol", "No, Yes",
"chol_check", "Factor", "Cholesterol check in past 5 years", "No, Yes",
"bmi", "Numeric", "Body Mass Index", "Continuous",
"smoker", "Factor", "Smoked 100+ cigarettes in life", "No, Yes",
"stroke", "Factor", "History of stroke", "No, Yes",
"heart_disease_or_attack", "Factor", "Coronary heart disease or MI", "No, Yes",
"phys_activity", "Factor", "Physical activity in past 30 days", "No, Yes",
"fruits", "Factor", "Consume fruit daily", "No, Yes",
"veggies", "Factor", "Consume vegetables daily", "No, Yes",
"hvy_alcohol_consump", "Factor", "Heavy alcohol use", "No, Yes",
"any_healthcare", "Factor", "Have healthcare coverage", "No, Yes",
"no_docbc_cost", "Factor", "Avoided doctor due to cost", "No, Yes",
"gen_hlth", "Numeric", "General health rating", "1–5",
"ment_hlth", "Numeric", "Poor mental health days", "0–30",
"phys_hlth", "Numeric", "Poor physical health days", "0–30",
"diff_walk", "Factor", "Difficulty walking", "No, Yes",
"sex", "Factor", "Biological sex", "Female, Male",
"age", "Numeric", "Age category", "1–13",
"education", "Numeric", "Education level", "1–6",
"income", "Numeric", "Income level", "1–8",
"bmi_category", "Factor", "BMI category", "Underweight–Obese",
"health_risk_score", "Numeric", "Composite health risk score", "0–7",
"lifestyle_score", "Numeric", "Healthy lifestyle score", "0–3"
)

write_csv(data_dictionary, "data/data_dictionary.csv")
```

```{r}
# Load processed data
diabetes <- read_csv("data/diabetes_processed.csv") %>%
  mutate(
    diabetes_binary = factor(diabetes_binary,
                             levels = c("No_Diabetes", "Diabetes"))
  )
```

# Research Questions

In this project we would like to focus on these three questions: 

1. Which demographic, lifestyle, and clinical characteristics are most strongly associated with diabetes?

2. How effectively can logistic regression models predict diabetes status using only self-reported survey information?

3. Does model complexity like interaction terms improve predictive performance?

# Data

We use the cleaned [2015 BRFSS dataset](https://www.kaggle.com/datasets/alexteboul/diabetes-health-indicators-dataset/data) (253,680 adults, 22 variables), focusing on variables related to demographics, lifestyle, and self-reported health.

- **Population:** Adults participating in the 2015 BRFSS survey.
- **Outcome:** Self-reported diabetes status (binary outcome used for prediction).
- **Main predictors:** BMI, age category, general health rating, blood pressure, cholesterol, smoking status, physical activity, sex, and socioeconomic indicators

## Key preprocessing steps:

- No missing data were present (already cleaned by BRFSS).

- Created interpretable new variables according to [dataset notebook](https://www.kaggle.com/code/alexteboul/diabetes-health-indicators-dataset-notebook) and [codebook](https://www.cdc.gov/brfss/annual_data/2015/pdf/codebook15_llcp.pdf). 

  - BMI categories

  - Mental & physical health categories

  - Composite lifestyle and health-risk scores

  - Converted categorical variables into meaningful factor labels

- Train/test Split

  - 80/20 stratified split is used to preserve diabetes prevalence.
  
  - Training set: approximately 202,944 observations (for EDA and model fitting)
  
  - Test set: approximately 50,736 observations (for final evaluation only)

- Fixed random seed (123) for reproducibility

## Exploratory analysis revealed:

- **Strong class imbalance (far more non-diabetic respondents).**

```{r}
# Distribution of Diabetes Status
diabetes %>%
  ggplot(aes(x = diabetes_binary, fill = diabetes_binary)) +
  geom_bar(width = 0.55, alpha = 0.9, color = "white") +
  scale_fill_manual(values = c("#5DA5DA", "#F17CB0")) + 
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(
    title = "Distribution of Diabetes Status",
    x = "Diabetes Status",
    y = "Count"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
    axis.title.x = element_text(size = 14, margin = margin(t = 10)),
    axis.title.y = element_text(size = 14, margin = margin(r = 10)),
    axis.text = element_text(size = 12),
    legend.position = "none",
    plot.margin = margin(15, 15, 15, 15)
  )

```

According to this plot, diabetes individual included is much less than non-diabetes. This may lead to strong class imbalance. 

```{r}
# BMI Histogram
diabetes %>%
  ggplot(aes(x = bmi)) +
  geom_histogram(bins = 40, fill = "#6495ED", alpha = 0.8) +
  labs(
    title = "Distribution of BMI",
    x = "BMI",
    y = "Count"
  ) +
  theme_minimal()
```

The BMI in this sample is right-skewed, with most values falling between 20 and 40. 

The concentration of observations in the overweight and obese range reflects the population distribution in this dataset. A smaller number of individuals have very high BMI values, forming a long right tail.

```{r}
# Age Category Distribution
diabetes %>%
  ggplot(aes(x = age_cat)) +
  geom_bar(fill = "#8B5FBF", alpha = 0.8) +
  labs(
    title = "Age Category Distribution",
    x = "Age Category",
    y = "Count"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Left-skewness in this plot shows that the dataset contains far more middle-aged and older adults than younger respondents. Counts increase steadily from the 18–24 group and peak between ages 55–69, before gradually declining in the 70+ categories. This uneven age distribution is important because diabetes risk rises with age. This means that the sample naturally contains more individuals from higher-risk age groups.

- **Diabetic individuals tend to have higher BMI, poor general health, and older age.**

```{r}
# BMI by Diabetes Status
diabetes %>%
  ggplot(aes(x = diabetes_binary, y = bmi, fill = diabetes_binary)) +
  geom_boxplot(alpha = 0.85) +
  scale_fill_manual(values = c("lightblue", "lightpink")) +
  labs(
    title = "BMI by Diabetes Status",
    x = "Diabetes Status",
    y = "BMI"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

The plot shows that individuals included with diabetes tend to have higher BMI values than those without diabetes. Both the median and overall distribution of diabetes group shifted upward. 

```{r}
# Age Category (Numeric) by Diabetes Status
diabetes %>%
  ggplot(aes(x = diabetes_binary, y = age, fill = diabetes_binary)) +
  geom_boxplot(alpha = 0.85) +
  labs(
    title = "Age Category by Diabetes Status",
    x = "Diabetes Status",
    y = "Age Category (1 = 18–24 … 13 = 80+)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

The plot shows that the median age category is higher for the diabetes group, and their overall distribution is shifted toward older age ranges.

```{r}
# General Health Score by Diabetes Status
diabetes %>%
  ggplot(aes(x = diabetes_binary, y = gen_hlth, fill = diabetes_binary)) +
  geom_boxplot(alpha = 0.85) +
  scale_fill_manual(values = c("#4C9F70", "#C94C4C")) +
  labs(
    title = "General Health Score by Diabetes Status",
    x = "Diabetes Status",
    y = "General Health (1 = Excellent, 5 = Poor)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

The plot shows that individuals with diabetes generally report worse overall health than those without diabetes. The diabetes group has a higher median health score, which indicates poorer health and a distribution shifted toward less favorable health ratings.

- **Income distribution is skewed toward higher brackets, which may influence access to care.**

```{r}
# Income Category Distribution
diabetes %>%
  ggplot(aes(x = income_cat)) +
  geom_bar(fill = "#FFA500", alpha = 0.8) +
  labs(
    title = "Income Level Distribution",
    x = "Income Category",
    y = "Count"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
# Correlation heat map
# Select numeric variables
numeric_vars <- diabetes %>%
  select(bmi, ment_hlth, phys_hlth, gen_hlth, income, age)

# Compute and tidy correlation matrix
cor_df <- numeric_vars %>%
  cor(use = "complete.obs") %>%
  as.data.frame() %>%
  rownames_to_column("var1") %>%
  pivot_longer(-var1, names_to = "var2", values_to = "cor")

# Heatmap with correlation values
cor_df %>%
  ggplot(aes(var1, var2, fill = cor)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(
    low = "#4575b4",
    high = "#d73027",
    mid = "white",
    midpoint = 0,
    limits = c(-1, 1)
  ) +
  geom_text(aes(label = round(cor, 2)), size = 3) +
  labs(
    title = "Correlation Heatmap of Numeric Predictors",
    x = "",
    y = "",
    fill = "Correlation"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The heatmap shows that correlations among the numeric predictors are generally weak to moderate, indicating low multicollinearity and supporting their joint use in logistic regression models. 

Variables related to overall health, such as general health and physical or mental health days, are moderately correlated as expected,

While BMI and age show only weak relationships with other predictors. 
Overall, the plot confirms that each variable contributes distinct information for diabetes risk prediction.

# Key Findings

## 1. Proposed Model Performed Best

We compared three logistic regression models:

- Main Effects Model

- BMI × Age Interaction Model

- High-Interaction Model

The Main Effects Model achieved the highest predictive performance (AUC = 0.806 vs. 0,799 vs. 0.781) and best calibration.

Adding interaction terms did not improve accuracy and sometimes reduced sensitivity, suggesting noise rather than signal was introduced.

```{r}
# Fit Logistic Regression Models
# Main Effects Model
logit_main <- glm(
diabetes_binary ~ bmi + age_cat + gen_hlth_cat + high_bp + high_chol +
phys_activity + smoker + heart_disease_or_attack + bmi_category +
income_cat + education_cat + sex,
data = diabetes_train,
family = binomial()
)
```

```{r}
# Subgroup Model: BMI × Age Category
logit_subgroup <- glm(
diabetes_binary ~ bmi * age_cat + gen_hlth_cat + high_bp +
phys_activity + smoker + income_cat + sex,
data = diabetes_train,
family = binomial()
)
```

```{r}
# High-Interaction Model
logit_interaction <- glm(
diabetes_binary ~ bmi * phys_activity * high_bp +
gen_hlth_cat + smoker + heart_disease_or_attack + sex,
data = diabetes_train,
family = binomial()
)
```

```{r}
# Model Performance Evaluation
# Utility Function
library(pROC)

evaluate_model <- function(model, data) {
  probs <- predict(model, newdata = data, type = "response")
  preds <- factor(ifelse(probs > 0.5, "Diabetes", "No_Diabetes"))
  truth <- data$diabetes_binary
  
  cm <- confusionMatrix(preds, truth, positive = "Diabetes")
  
  tibble(
    Accuracy    = cm$overall["Accuracy"],
    Sensitivity = cm$byClass["Sensitivity"],
    Specificity = cm$byClass["Specificity"],
    AUC         = as.numeric(pROC::auc(truth, probs)) 
  )
}

```

```{r}
# Compute Test Set Performance
results_main <- evaluate_model(logit_main, diabetes_test)
results_subgroup <- evaluate_model(logit_subgroup, diabetes_test)
results_interaction <- evaluate_model(logit_interaction, diabetes_test)

results_all <- bind_rows(
"Main Effects" = results_main,
"BMI × Age Subgroup" = results_subgroup,
"Interaction Model" = results_interaction,
.id = "Model"
)
```

```{r}
# ROC Curves for All Three Models
probs_main <- predict(logit_main, diabetes_test, type = "response")
probs_sub  <- predict(logit_subgroup, diabetes_test, type = "response")
probs_int  <- predict(logit_interaction, diabetes_test, type = "response")

roc_main <- roc(diabetes_test$diabetes_binary, probs_main)
roc_sub  <- roc(diabetes_test$diabetes_binary, probs_sub)
roc_int  <- roc(diabetes_test$diabetes_binary, probs_int)
```

```{r}
# Variable Importance (Absolute Log-Odds Effect)
var_imp <- function(model){
broom::tidy(model) %>%
filter(term != "(Intercept)") %>%
mutate(abs_effect = abs(estimate)) %>%
arrange(desc(abs_effect))
}
```

```{r, include = FALSE}
# Final Model Comparison Table
results_all %>%
mutate(across(where(is.numeric), round, 3))
```

```{r}
## Predictive model evaluation (ROC/AUC, confusion matrix, feature importance)
# 1. Generate Probabilities for All Models
# -------------------------------------------------------
probs_main <- predict(logit_main, diabetes_test, type = "response")
probs_sub  <- predict(logit_subgroup, diabetes_test, type = "response")
probs_int  <- predict(logit_interaction, diabetes_test, type = "response")
```

```{r}
# Calculate ROC objects
roc_main <- roc(diabetes_test$diabetes_binary, probs_main)
roc_sub  <- roc(diabetes_test$diabetes_binary, probs_sub)
roc_int  <- roc(diabetes_test$diabetes_binary, probs_int)

# Plot ROC Curves
ggroc(list("Main Effects" = roc_main, 
           "BMI × Age" = roc_sub, 
           "Interaction" = roc_int)) +
  scale_color_manual(values = c("#1f77b4", "#ff7f0e", "#2ca02c")) +
  geom_abline(intercept = 0, slope = 1, color = "gray", linetype = "dashed") +
  labs(title = "ROC Curves Comparison",
       x = "Specificity",
       y = "Sensitivity",
       color = "Model") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom")

# Threshold Tuning
best_coords <- coords(roc_main, "best", best.method = "youden", 
                      ret = c("threshold", "specificity", "sensitivity"))

# Extract the numeric threshold safely (taking the first if there are ties)
optimal_threshold <- as.numeric(best_coords$threshold[1])

# Generate predictions using this new threshold
preds_tuned <- factor(ifelse(probs_main > optimal_threshold, "Diabetes", "No_Diabetes"),
                      levels = c("No_Diabetes", "Diabetes"))
```

The ROC curves show that the Main Effects model consistently achieves the highest sensitivity across specificity levels. This confirms its better discrimination ability compared with the BMI × Age and High-Interaction models. 

The two more complex models both produce slightly lower ROC curves. Additional interaction terms introduced noise and did not improve predictive performance. 

The simpler Main Effects model is the most effective and reliable choice for diabetes risk prediction.

## 2. Significant Predictors of Diabetes

The variables most strongly associated with diabetes were:

- age level
- general health rating
- BMI
- high blood pressure
- high cholesterol
- socioeconomic indicators

```{r}
# Main Effects Model

tidy(logit_main) %>%
  filter(term != "(Intercept)") %>%
  mutate(abs_effect = abs(estimate),
         term = str_remove_all(term, "cat|factor|\\(|\\)")) %>% 
  slice_max(order_by = abs_effect, n = 10) %>%
  ggplot(aes(x = abs_effect, y = reorder(term, abs_effect))) +
  geom_col(fill = "#4C9F70", width = 0.7) +
  labs(title = "Top 10 Predictors of Diabetes Risk",
       subtitle = "Based on Absolute Log-Odds",
       x = "Effect Size (Absolute Log-Odds)",
       y = NULL) +   
  theme_minimal(base_size = 14) 
```

These findings align with established clinical literature, which means they support the external validity of our model.

## 3. Importance of Adjusting the Classification Threshold

Diabetes is relatively rare in this dataset. The default probability cutoff of 0.5 produced **high accuracy (~85%) and dangerously low sensitivity (~15%)**. 

Adjusting the threshold using **Youden’s Index (~0.14)** significantly **improved sensitivity** while maintaining acceptable specificity. This improved the model in ability of identifying high-risk individuals from public health screening.

```{r}
# Calculate confusion matrix stats
cm_tuned <- confusionMatrix(preds_tuned, diabetes_test$diabetes_binary, positive = "Diabetes")

# Visualizing the Confusion Matrix as a Heatmap
cm_data <- as.data.frame(cm_tuned$table)

ggplot(cm_data, aes(x = Prediction, y = Reference, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), size = 6, color = "white") +
  scale_fill_gradient(low = "#8ebad9", high = "#0b406c") +
  labs(title = paste("Confusion Matrix (Threshold =", round(optimal_threshold, 3), ")"),
       subtitle = "Optimized for Screening (Higher Sensitivity)",
       x = "Predicted Label",
       y = "Actual Label") +
  theme_minimal(base_size = 14)
```

This confusion matrix illustrates the performance of the Main Effects model after lowering the decision threshold to 0.141. 

With this threshold, the model correctly identifies far more true diabetes cases (1,480) than at the default threshold, though at the cost of increasing false positives (12,467). 

This trade-off is more appropriate for screening, as missing true cases is more harmful than referring additional individuals for follow-up testing.

## 4. Takeaway Messages

According to our model, predictors from a short list of survey questions that are included in our model, especially **age, general health rating, BMI**, may be able to provide reliable early risk stratification without clinical testing.

**A straightforward model using basic survey questions can still pick out individuals who are at higher risk for diabetes reliably. Making the current model more complicated didn’t improve the model.**

**This means early diabetes screening could be done with quick survey data before medical tests, which makes the screening more affordable and accessible.**
