---
title: "Task 1: Data Cleaning and Processing"
author: "cl4775"
date: "December 2025"
output: github_document
---

# P8105 Final Project: Diabetes Health Indicators
 
**Task 1: Data Cleaning, Variable Processing, and Train/Test Dataset Creation**

Dataset: CDC BRFSS 2015 Diabetes Health Indicators  
Source: https://www.kaggle.com/datasets/alexteboul/diabetes-health-indicators-dataset

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Load Required Libraries

```{r load_libraries}
library(tidyverse)
library(caret)
library(skimr)
library(janitor)
library(corrplot)
library(DataExplorer)

# Set seed for reproducibility
set.seed(123)
```

## 1. Data Import

The dataset has three versions:
- `diabetes_012_health_indicators_BRFSS2015.csv` (3 classes: 0, 1, 2)
- `diabetes_binary_5050split_health_indicators_BRFSS2015.csv` (balanced binary)
- `diabetes_binary_health_indicators_BRFSS2015.csv` (full binary)

For this project, we'll use the full binary version for more data.

```{r import_data}
diabetes_raw <- read_csv("diabetes_binary_health_indicators_BRFSS2015.csv")

glimpse(diabetes_raw)
```

## 2. Initial Data Exploration

```{r explore_data}
# Check for missing values
colSums(is.na(diabetes_raw))

# Summary statistics
summary(diabetes_raw)

# Check for duplicates
sum(duplicated(diabetes_raw))

# Examine outcome variable distribution
table(diabetes_raw$Diabetes_binary)
prop.table(table(diabetes_raw$Diabetes_binary))
```

## 3. Data Cleaning

```{r clean_data}
diabetes_clean <- diabetes_raw %>%
  # Clean column names
  clean_names() %>%
  
  # Fix the heart disease variable name
  rename(heart_disease_or_attack = heart_diseaseor_attack) %>%
  
  # Remove duplicates if any
  distinct() %>%
  
  # Remove rows with missing values (if any)
  drop_na()

glimpse(diabetes_clean)
```

## 4. Variable Processing and Feature Engineering

### Variable Descriptions

According to the dataset documentation:

1. **Diabetes_binary**: 0 = no diabetes, 1 = prediabetes or diabetes
2. **HighBP**: 0 = no, 1 = yes
3. **HighChol**: 0 = no, 1 = yes
4. **CholCheck**: 0 = no, 1 = yes (cholesterol check in 5 years)
5. **BMI**: Body Mass Index
6. **Smoker**: 0 = no, 1 = yes
7. **Stroke**: 0 = no, 1 = yes
8. **HeartDiseaseorAttack**: 0 = no, 1 = yes
9. **PhysActivity**: 0 = no, 1 = yes
10. **Fruits**: 0 = no, 1 = yes (consume fruit 1+ times per day)
11. **Veggies**: 0 = no, 1 = yes (consume vegetables 1+ times per day)
12. **HvyAlcoholConsump**: 0 = no, 1 = yes
13. **AnyHealthcare**: 0 = no, 1 = yes
14. **NoDocbcCost**: 0 = no, 1 = yes (couldn't see doctor due to cost)
15. **GenHlth**: 1-5 scale (1 = excellent, 5 = poor)
16. **MentHlth**: days of poor mental health in past 30 days (0-30)
17. **PhysHlth**: days of poor physical health in past 30 days (0-30)
18. **DiffWalk**: 0 = no, 1 = yes (difficulty walking or climbing stairs)
19. **Sex**: 0 = female, 1 = male
20. **Age**: 1-13 (age categories)
21. **Education**: 1-6 (education level)
22. **Income**: 1-8 (income categories)

### Processing

```{r process_variables}
diabetes_processed <- diabetes_clean %>%
  mutate(
    # Convert outcome to factor
    diabetes_binary = factor(diabetes_binary, 
                             levels = c(0, 1),
                             labels = c("No_Diabetes", "Diabetes")),
    
    # Convert binary variables to factors
    high_bp = factor(high_bp, levels = c(0, 1), labels = c("No", "Yes")),
    high_chol = factor(high_chol, levels = c(0, 1), labels = c("No", "Yes")),
    chol_check = factor(chol_check, levels = c(0, 1), labels = c("No", "Yes")),
    smoker = factor(smoker, levels = c(0, 1), labels = c("No", "Yes")),
    stroke = factor(stroke, levels = c(0, 1), labels = c("No", "Yes")),
    heart_disease_or_attack = factor(heart_disease_or_attack, 
                                     levels = c(0, 1), 
                                     labels = c("No", "Yes")),
    phys_activity = factor(phys_activity, levels = c(0, 1), labels = c("No", "Yes")),
    fruits = factor(fruits, levels = c(0, 1), labels = c("No", "Yes")),
    veggies = factor(veggies, levels = c(0, 1), labels = c("No", "Yes")),
    hvy_alcohol_consump = factor(hvy_alcohol_consump, 
                                 levels = c(0, 1), 
                                 labels = c("No", "Yes")),
    any_healthcare = factor(any_healthcare, levels = c(0, 1), labels = c("No", "Yes")),
    no_docbc_cost = factor(no_docbc_cost, levels = c(0, 1), labels = c("No", "Yes")),
    diff_walk = factor(diff_walk, levels = c(0, 1), labels = c("No", "Yes")),
    sex = factor(sex, levels = c(0, 1), labels = c("Female", "Male")),
    
    # Convert ordinal variables to factors
    gen_hlth_cat = factor(gen_hlth, 
                          levels = 1:5,
                          labels = c("Excellent", "Very Good", "Good", "Fair", "Poor"),
                          ordered = TRUE),
    
    # Create age groups
    age_cat = factor(age,
                     levels = 1:13,
                     labels = c("18-24", "25-29", "30-34", "35-39", "40-44",
                                "45-49", "50-54", "55-59", "60-64", "65-69",
                                "70-74", "75-79", "80+"),
                     ordered = TRUE),
    
    # Create education categories
    education_cat = factor(education,
                           levels = 1:6,
                           labels = c("Never/K-8", "Grades 9-11", "Grade 12/GED",
                                      "College 1-3 yrs", "College 4+ yrs", "Unknown"),
                           ordered = TRUE),
    
    # Create income categories
    income_cat = factor(income,
                        levels = 1:8,
                        labels = c("<$10k", "$10k-$15k", "$15k-$20k", "$20k-$25k",
                                   "$25k-$35k", "$35k-$50k", "$50k-$75k", "$75k+"),
                        ordered = TRUE),
    
    # Create BMI categories
    bmi_category = case_when(
      bmi < 18.5 ~ "Underweight",
      bmi >= 18.5 & bmi < 25 ~ "Normal",
      bmi >= 25 & bmi < 30 ~ "Overweight",
      bmi >= 30 ~ "Obese"
    ) %>% factor(levels = c("Underweight", "Normal", "Overweight", "Obese"),
                 ordered = TRUE),
    
    # Create mental health categories
    ment_hlth_cat = case_when(
      ment_hlth == 0 ~ "None",
      ment_hlth <= 7 ~ "1-7 days",
      ment_hlth <= 14 ~ "8-14 days",
      ment_hlth <= 30 ~ "15-30 days"
    ) %>% factor(levels = c("None", "1-7 days", "8-14 days", "15-30 days"),
                 ordered = TRUE),
    
    # Create physical health categories
    phys_hlth_cat = case_when(
      phys_hlth == 0 ~ "None",
      phys_hlth <= 7 ~ "1-7 days",
      phys_hlth <= 14 ~ "8-14 days",
      phys_hlth <= 30 ~ "15-30 days"
    ) %>% factor(levels = c("None", "1-7 days", "8-14 days", "15-30 days"),
                 ordered = TRUE),
    
    # Create composite health score
    health_risk_score = as.numeric(high_bp == "Yes") +
      as.numeric(high_chol == "Yes") +
      as.numeric(stroke == "Yes") +
      as.numeric(heart_disease_or_attack == "Yes") +
      as.numeric(smoker == "Yes") +
      as.numeric(hvy_alcohol_consump == "Yes") +
      (gen_hlth - 1) / 4,
    
    # Create healthy lifestyle score
    lifestyle_score = as.numeric(phys_activity == "Yes") +
      as.numeric(fruits == "Yes") +
      as.numeric(veggies == "Yes")
  )

glimpse(diabetes_processed)
```

## 5. Data Quality Checks

```{r quality_checks}
# Check for any issues in processed data
sum(is.na(diabetes_processed))

# Verify factor levels
table(diabetes_processed$diabetes_binary)

# Check continuous variables for outliers
diabetes_processed %>%
  select(bmi, ment_hlth, phys_hlth, health_risk_score, lifestyle_score) %>%
  summary()
```

## 6. Train/Test Split

Create stratified train/test split (80/20) using stratified sampling to maintain outcome distribution.

```{r train_test_split}
# Set seed again for reproducibility
set.seed(123)

# Create partition index
train_index <- createDataPartition(
  diabetes_processed$diabetes_binary,
  p = 0.80,
  list = FALSE,
  times = 1
)

# Create training and testing datasets
diabetes_train <- diabetes_processed[train_index, ]
diabetes_test <- diabetes_processed[-train_index, ]

# Verify split proportions
nrow(diabetes_train)
nrow(diabetes_test)

# Check outcome distribution in train and test sets
prop.table(table(diabetes_train$diabetes_binary))
prop.table(table(diabetes_test$diabetes_binary))
```

## 7. Export Cleaned Datasets

```{r export_data}
# Create output directory if it doesn't exist
if (!dir.exists("data")) {
  dir.create("data")
}

# Save processed datasets
write_csv(diabetes_processed, "data/diabetes_processed.csv")
write_csv(diabetes_train, "data/diabetes_train.csv")
write_csv(diabetes_test, "data/diabetes_test.csv")

# Also save as RDS for faster loading in R
saveRDS(diabetes_processed, "data/diabetes_processed.rds")
saveRDS(diabetes_train, "data/diabetes_train.rds")
saveRDS(diabetes_test, "data/diabetes_test.rds")
```

## 8. Generate Data Dictionary

```{r data_dictionary}
# Create data dictionary
data_dictionary <- tribble(
  ~Variable, ~Type, ~Description, ~Values,
  "diabetes_binary", "Factor", "Diabetes diagnosis", "No_Diabetes, Diabetes",
  "high_bp", "Factor", "High blood pressure", "No, Yes",
  "high_chol", "Factor", "High cholesterol", "No, Yes",
  "chol_check", "Factor", "Cholesterol check in past 5 years", "No, Yes",
  "bmi", "Numeric", "Body Mass Index", "Continuous",
  "smoker", "Factor", "Smoked at least 100 cigarettes in life", "No, Yes",
  "stroke", "Factor", "History of stroke", "No, Yes",
  "heart_disease_or_attack", "Factor", "Coronary heart disease or MI", "No, Yes",
  "phys_activity", "Factor", "Physical activity in past 30 days", "No, Yes",
  "fruits", "Factor", "Consume fruit 1+ times per day", "No, Yes",
  "veggies", "Factor", "Consume vegetables 1+ times per day", "No, Yes",
  "hvy_alcohol_consump", "Factor", "Heavy alcohol consumption", "No, Yes",
  "any_healthcare", "Factor", "Have any healthcare coverage", "No, Yes",
  "no_docbc_cost", "Factor", "Could not see doctor due to cost", "No, Yes",
  "gen_hlth", "Numeric", "General health rating", "1-5 (1=excellent, 5=poor)",
  "ment_hlth", "Numeric", "Days of poor mental health (past 30)", "0-30",
  "phys_hlth", "Numeric", "Days of poor physical health (past 30)", "0-30",
  "diff_walk", "Factor", "Difficulty walking or climbing stairs", "No, Yes",
  "sex", "Factor", "Biological sex", "Female, Male",
  "age", "Numeric", "Age category", "1-13 (see age_cat for labels)",
  "education", "Numeric", "Education level", "1-6",
  "income", "Numeric", "Income level", "1-8",
  "bmi_category", "Factor", "BMI category", "Underweight, Normal, Overweight, Obese",
  "health_risk_score", "Numeric", "Composite health risk score", "Continuous (0-7)",
  "lifestyle_score", "Numeric", "Healthy lifestyle score", "0-3"
)

# Save data dictionary
write_csv(data_dictionary, "data/data_dictionary.csv")

# Display dictionary
data_dictionary
```
