---
title: "Final Project"
author: "Yuwei Ma, Rachel Lu, Chenxi Liu, Xiuhong Fan, Daisy Gui"
date: "December 2025"
output: github_document
---



## Exploratory analysis

```{r}
# Load processed data
diabetes <- read_csv("data/diabetes_processed.csv") %>%
  mutate(
    diabetes_binary = factor(diabetes_binary,
                             levels = c("No_Diabetes", "Diabetes"))
  )
```

# Research Questions

In this project we would like to focus on these three questions: 

1. Which demographic, lifestyle, and clinical characteristics are most strongly associated with diabetes?

2. How effectively can logistic regression models predict diabetes status using only self-reported survey information?

3. Does model complexity like interaction terms improve predictive performance?

# Data

We use the cleaned [2015 BRFSS dataset](https://www.kaggle.com/datasets/alexteboul/diabetes-health-indicators-dataset/data) (253,680 adults, 22 variables), focusing on variables related to demographics, lifestyle, and self-reported health.

- **Population:** Adults participating in the 2015 BRFSS survey.
- **Outcome:** Self-reported diabetes status (binary outcome used for prediction).
- **Main predictors:** BMI, age category, general health rating, blood pressure, cholesterol, smoking status, physical activity, sex, and socioeconomic indicators

## Key preprocessing steps:

- No missing data were present (already cleaned by BRFSS).

- Created interpretable new variables according to [dataset notebook](https://www.kaggle.com/code/alexteboul/diabetes-health-indicators-dataset-notebook) and [codebook](https://www.cdc.gov/brfss/annual_data/2015/pdf/codebook15_llcp.pdf). 

  - BMI categories

  - Mental & physical health categories

  - Composite lifestyle and health-risk scores

  - Converted categorical variables into meaningful factor labels

- Train/test Split

  - 80/20 stratified split is used to preserve diabetes prevalence.
  
  - Training set: approximately 202,944 observations (for EDA and model fitting)
  
  - Test set: approximately 50,736 observations (for final evaluation only)

- Fixed random seed (123) for reproducibility

## Exploratory analysis revealed:

- **Strong class imbalance (far more non-diabetic respondents).**

```{r}
# Distribution of Diabetes Status
diabetes %>%
  ggplot(aes(x = diabetes_binary, fill = diabetes_binary)) +
  geom_bar(width = 0.55, alpha = 0.9, color = "white") +
  scale_fill_manual(values = c("#5DA5DA", "#F17CB0")) + 
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(
    title = "Distribution of Diabetes Status",
    x = "Diabetes Status",
    y = "Count"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
    axis.title.x = element_text(size = 14, margin = margin(t = 10)),
    axis.title.y = element_text(size = 14, margin = margin(r = 10)),
    axis.text = element_text(size = 12),
    legend.position = "none",
    plot.margin = margin(15, 15, 15, 15)
  )

```

According to this plot, diabetes individual included is much less than non-diabetes. This may lead to strong class imbalance. 

```{r}
# BMI Histogram
diabetes %>%
  ggplot(aes(x = bmi)) +
  geom_histogram(bins = 40, fill = "#6495ED", alpha = 0.8) +
  labs(
    title = "Distribution of BMI",
    x = "BMI",
    y = "Count"
  ) +
  theme_minimal()
```

The BMI in this sample is right-skewed, with most values falling between 20 and 40. 

The concentration of observations in the overweight and obese range reflects the population distribution in this dataset. A smaller number of individuals have very high BMI values, forming a long right tail.

```{r}
# Age Category Distribution
diabetes %>%
  ggplot(aes(x = age_cat)) +
  geom_bar(fill = "#8B5FBF", alpha = 0.8) +
  labs(
    title = "Age Category Distribution",
    x = "Age Category",
    y = "Count"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Left-skewness in this plot shows that the dataset contains far more middle-aged and older adults than younger respondents. Counts increase steadily from the 18–24 group and peak between ages 55–69, before gradually declining in the 70+ categories. This uneven age distribution is important because diabetes risk rises with age. This means that the sample naturally contains more individuals from higher-risk age groups.

- **Diabetic individuals tend to have higher BMI, poor general health, and older age.**

```{r}
# BMI by Diabetes Status
diabetes %>%
  ggplot(aes(x = diabetes_binary, y = bmi, fill = diabetes_binary)) +
  geom_boxplot(alpha = 0.85) +
  scale_fill_manual(values = c("lightblue", "lightpink")) +
  labs(
    title = "BMI by Diabetes Status",
    x = "Diabetes Status",
    y = "BMI"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

The plot shows that individuals included with diabetes tend to have higher BMI values than those without diabetes. Both the median and overall distribution of diabetes group shifted upward. 

```{r}
# Age Category (Numeric) by Diabetes Status
diabetes %>%
  ggplot(aes(x = diabetes_binary, y = age, fill = diabetes_binary)) +
  geom_boxplot(alpha = 0.85) +
  labs(
    title = "Age Category by Diabetes Status",
    x = "Diabetes Status",
    y = "Age Category (1 = 18–24 … 13 = 80+)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

The plot shows that the median age category is higher for the diabetes group, and their overall distribution is shifted toward older age ranges.

```{r}
# General Health Score by Diabetes Status
diabetes %>%
  ggplot(aes(x = diabetes_binary, y = gen_hlth, fill = diabetes_binary)) +
  geom_boxplot(alpha = 0.85) +
  scale_fill_manual(values = c("#4C9F70", "#C94C4C")) +
  labs(
    title = "General Health Score by Diabetes Status",
    x = "Diabetes Status",
    y = "General Health (1 = Excellent, 5 = Poor)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

The plot shows that individuals with diabetes generally report worse overall health than those without diabetes. The diabetes group has a higher median health score, which indicates poorer health and a distribution shifted toward less favorable health ratings.

- **Income distribution is skewed toward higher brackets, which may influence access to care.**

```{r}
# Income Category Distribution
diabetes %>%
  ggplot(aes(x = income_cat)) +
  geom_bar(fill = "#FFA500", alpha = 0.8) +
  labs(
    title = "Income Level Distribution",
    x = "Income Category",
    y = "Count"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
# Correlation heat map
# Select numeric variables
numeric_vars <- diabetes %>%
  select(bmi, ment_hlth, phys_hlth, gen_hlth, income, age)

# Compute and tidy correlation matrix
cor_df <- numeric_vars %>%
  cor(use = "complete.obs") %>%
  as.data.frame() %>%
  rownames_to_column("var1") %>%
  pivot_longer(-var1, names_to = "var2", values_to = "cor")

# Heatmap with correlation values
cor_df %>%
  ggplot(aes(var1, var2, fill = cor)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(
    low = "#4575b4",
    high = "#d73027",
    mid = "white",
    midpoint = 0,
    limits = c(-1, 1)
  ) +
  geom_text(aes(label = round(cor, 2)), size = 3) +
  labs(
    title = "Correlation Heatmap of Numeric Predictors",
    x = "",
    y = "",
    fill = "Correlation"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The heatmap shows that correlations among the numeric predictors are generally weak to moderate, indicating low multicollinearity and supporting their joint use in logistic regression models. 

Variables related to overall health, such as general health and physical or mental health days, are moderately correlated as expected,

While BMI and age show only weak relationships with other predictors. 
Overall, the plot confirms that each variable contributes distinct information for diabetes risk prediction.


## Additional analysis

# Logistic regression modeling (main effects, subgroups, interactions) 

# Fit Logistic Regression Models
```{r}
diabetes_train_model <- diabetes_train %>%
  mutate(
    bmi_c = bmi - mean(bmi, na.rm = TRUE),

    age_cat        = fct_relevel(age_cat, "18-24"),
    gen_hlth_cat   = fct_relevel(gen_hlth_cat, "Excellent"),
    bmi_category   = fct_relevel(bmi_category, "Normal"),
    income_cat     = fct_relevel(income_cat, "$75k+"),
    education_cat  = fct_relevel(education_cat, "College 4+ yrs"),
    sex            = fct_relevel(sex, "Female"),
    high_bp        = fct_relevel(high_bp, "No"),
    high_chol      = fct_relevel(high_chol, "No"),
    phys_activity  = fct_relevel(phys_activity, "No"),
    smoker         = fct_relevel(smoker, "No"),
    heart_disease_or_attack = fct_relevel(heart_disease_or_attack, "No")
  )
```

## Main Effects Model
```{r}
logit_main <- glm(
diabetes_binary ~ bmi + age_cat + gen_hlth_cat + high_bp + high_chol +
phys_activity + smoker + heart_disease_or_attack + bmi_category +
income_cat + education_cat + sex,
data = diabetes_train,
family = binomial()
)

summary(logit_main)
```

BMI, poor self-rated health, high blood pressure, and high cholesterol were independently associated with higher odds of diabetes. Each 1-unit increase in BMI raised the odds by approximately 3–4%, while high blood pressure nearly doubled the odds. Poor general health demonstrated the strongest association, with progressively higher odds across worsening categories. Male sex was associated with a modest increase in diabetes risk.

## Subgroup Model: BMI × Age Category
```{r}
logit_subgroup <- glm(
diabetes_binary ~ bmi * age_cat + gen_hlth_cat + high_bp +
phys_activity + smoker + income_cat + sex,
data = diabetes_train,
family = binomial()
)

summary(logit_subgroup)
```

BMI remained a significant predictor across all age groups, with evidence of a slightly stronger association in older adults. Interaction terms indicated modest age-related variation in the BMI–diabetes relationship, consistent with heightened metabolic vulnerability in later life.

```{r}
logit_interaction <- glm(
diabetes_binary ~ bmi * phys_activity * high_bp +
gen_hlth_cat + smoker + heart_disease_or_attack + sex,
data = diabetes_train,
family = binomial()
)

summary(logit_interaction)
```

BMI and high blood pressure remained strong predictors, but neither the two-way nor three-way interaction terms materially improved model fit. There was no evidence that the combined effects of BMI, physical activity, and hypertension exceeded their independent contributions.

## Nonlinear BMI Effects (Restricted Cubic Splines)
```{r}
library(splines)

logit_spline <- glm(
  diabetes_binary ~ ns(bmi, df = 4) +
    age_cat + gen_hlth_cat + high_bp + smoker + income_cat + sex,
  data = diabetes_train_model,
  family = binomial()
)

summary(logit_spline)
bmi_range <- seq(min(diabetes_train$bmi), max(diabetes_train$bmi), length = 100)

pred_df <- tibble(
  bmi = bmi_range,
  age_cat = "18-24",
  gen_hlth_cat = "Excellent",
  high_bp = "No",
  smoker = "No",
  income_cat = "$75k+",
  sex = "Female"
)

pred_df$pred <- predict(logit_spline, newdata = pred_df, type = "response")

pred_df %>%
  ggplot(aes(bmi, pred)) +
  geom_line(color = "blue") +
  labs(title = "Nonlinear BMI–Diabetes Risk Curve (Spline Model)")
```

Restricted cubic splines revealed a nonlinear BMI–diabetes relationship, with risk rising gradually at lower BMI values and increasing more steeply in the overweight and obese ranges. This pattern indicates accelerating risk with increasing adiposity.

## Sex-Based Effect Modification (BMI × Sex)
```{r}
logit_bmi_sex <- glm(
  diabetes_binary ~ bmi_c * sex + age_cat + high_bp + gen_hlth_cat,
  data = diabetes_train_model,
  family = binomial()
)

summary(logit_bmi_sex)
coef(logit_bmi_sex)[grep("bmi_c", names(coef(logit_bmi_sex)))]
```

BMI was positively associated with diabetes in both sexes, with only minimal evidence of sex-based effect modification. Men exhibited higher baseline odds of diabetes, but the BMI slope differed only slightly between sexes.

## Income × General Health (Health Disparity Model)
```{r}
logit_income_health <- glm(
  diabetes_binary ~ income_cat * gen_hlth_cat +
    age_cat + bmi_c + high_bp + high_chol,
  data = diabetes_train_model,
  family = binomial()
)

summary(logit_income_health)
```

Lower income and poorer self-rated health were independently associated with elevated diabetes risk; however, no meaningful interaction was observed. The effects of income and health status appeared additive rather than synergistic.

## Psychological Health × Physical Activity Interaction
```{r}
logit_mental_activity <- glm(
  diabetes_binary ~ ment_hlth_cat * phys_activity +
    bmi_c + age_cat + high_bp + high_chol + smoker,
  data = diabetes_train_model,
  family = binomial()
)

summary(logit_mental_activity)
```

Poor mental health and physical inactivity were each associated with increased diabetes risk, but interaction terms were not significant. Physical activity did not substantially modify the relationship between mental health burden and diabetes.

## Multi-Morbidity (Health Risk Score Model)
```{r}
logit_multi_morbidity <- glm(
  diabetes_binary ~ health_risk_score + age_cat + income_cat + sex,
  data = diabetes_train_model,
  family = binomial()
)

summary(logit_multi_morbidity)
```

The health risk score showed a strong dose–response relationship, with each additional comorbidity substantially increasing odds of diabetes. This summary metric efficiently captured cumulative cardiometabolic burden.

## Lifestyle Score Model (Fruits + Veggies + Physical Activity)
```{r}
logit_lifestyle <- glm(
  diabetes_binary ~ lifestyle_score + age_cat + bmi_c +
    high_bp + gen_hlth_cat + income_cat,
  data = diabetes_train_model,
  family = binomial()
)

summary(logit_lifestyle)
```

Higher lifestyle scores—reflecting healthier diet and physical activity—were modestly protective, though the effect size was small relative to BMI and high blood pressure. Lifestyle behaviors contributed independently to risk reduction.

## Stratified Regression (True Stratification)
```{r}
young_group <- diabetes_train_model %>%
  filter(age_cat %in% c("18-24","25-29","30-34"))

old_group <- diabetes_train_model %>%
  filter(age_cat %in% c("60-64","65-69","70-74","75-79","80+"))
logit_young <- glm(
  diabetes_binary ~ bmi_c + gen_hlth_cat + high_bp,
  data = young_group, family = binomial()
)

logit_old <- glm(
  diabetes_binary ~ bmi_c + gen_hlth_cat + high_bp,
  data = old_group, family = binomial()
)

summary(logit_young)
summary(logit_old)
```

In stratified analyses, BMI and high blood pressure were important predictors in both age groups, but associations were stronger in older adults. Poor self-rated health exhibited pronounced effects across strata, with particularly elevated odds in younger individuals.

## A Reduced, Clinically Interpretable Core Model
```{r}
logit_core <- glm(
  diabetes_binary ~ bmi_c + age_cat + high_bp + gen_hlth_cat + sex,
  data = diabetes_train_model,
  family = binomial()
)

summary(logit_core)
```
## Final Regression Table Code (JAMA / NEJM Style)
```{r}
library(broom)
library(dplyr)
library(knitr)

core_tbl <- tidy(logit_core, conf.int = TRUE, exponentiate = TRUE) %>%
  mutate(
    OR = round(estimate, 2),
    CI_low = round(conf.low, 2),
    CI_high = round(conf.high, 2),
    OR_CI = paste0(OR, " (", CI_low, ", ", CI_high, ")"),
    p_value = ifelse(p.value < 0.001, "<0.001", sprintf("%.3f", p.value))
  ) %>%
  mutate(
    term = recode(term,
      "(Intercept)" = "Intercept",
      "bmi_c" = "BMI (per 1-unit increase)",
      "sexMale" = "Male (vs Female)",
      "high_bpYes" = "High blood pressure (Yes vs No)",

      # Age categories
      "age_cat25-29" = "Age 25–29 (vs 18–24)",
      "age_cat30-34" = "Age 30–34",
      "age_cat35-39" = "Age 35–39",
      "age_cat40-44" = "Age 40–44",
      "age_cat45-49" = "Age 45–49",
      "age_cat50-54" = "Age 50–54",
      "age_cat55-59" = "Age 55–59",
      "age_cat60-64" = "Age 60–64",
      "age_cat65-69" = "Age 65–69",
      "age_cat70-74" = "Age 70–74",
      "age_cat75-79" = "Age 75–79",
      "age_cat80+"   = "Age 80+",

      # General health
      "gen_hlth_catVery Good" = "Very good health (vs Excellent)",
      "gen_hlth_catGood"      = "Good health",
      "gen_hlth_catFair"      = "Fair health",
      "gen_hlth_catPoor"      = "Poor health"
    )
  ) %>%
  select(
    Predictor = term,
    OR,
    `95% CI` = OR_CI,
    p_value
  )

kable(
  core_tbl,
  caption = "Final Multivariable Logistic Regression Model for Diabetes (JAMA/NEJM Style)",
  col.names = c("Predictor", "Adjusted OR", "95% CI", "P Value"),
  align = "lccc"
)
```

In the final multivariable model, several predictors demonstrated strong and independent associations with diabetes. Higher BMI was associated with increased risk, with each 1-unit increase corresponding to a 6% higher odds of diabetes (adjusted OR 1.06, 95% CI 1.06–1.06). Age showed a pronounced nonlinear pattern: the dominant linear contrast was strongly positive (OR 16.10, 95% CI 13.96–18.69), while higher-order contrasts were less than 1, reflecting curvature in the age–risk relationship. High blood pressure was among the strongest clinical predictors, associated with more than double the odds of diabetes (OR 2.30, 95% CI 2.23–2.37). Self-reported general health also showed a large gradient, with the primary contrast indicating a more than six-fold increase in diabetes risk for those reporting worse health (OR 6.27, 95% CI 5.94–6.61). Male sex was associated with moderately elevated odds (OR 1.26, 95% CI 1.22–1.29). Together, these findings highlight the dominant role of adiposity, age-related risk, blood pressure, and overall perceived health status in determining diabetes risk.

# Predictive model evaluation (ROC/AUC, confusion matrix, feature importance)
```{r}
# 1. Generate Probabilities for All Models
# -------------------------------------------------------
probs_main <- predict(logit_main, diabetes_test, type = "response")
probs_sub  <- predict(logit_subgroup, diabetes_test, type = "response")
probs_int  <- predict(logit_interaction, diabetes_test, type = "response")
```


```{r}
# Calculate ROC objects
roc_main <- roc(diabetes_test$diabetes_binary, probs_main)
roc_sub  <- roc(diabetes_test$diabetes_binary, probs_sub)
roc_int  <- roc(diabetes_test$diabetes_binary, probs_int)

# Print AUC values
print(paste("AUC Main Effects:", auc(roc_main)))
print(paste("AUC Subgroup:", auc(roc_sub)))
print(paste("AUC Interaction:", auc(roc_int)))

# Plot ROC Curves (using ggplot2 for better visuals)
ggroc(list("Main Effects" = roc_main, 
           "BMI × Age" = roc_sub, 
           "Interaction" = roc_int)) +
  scale_color_manual(values = c("#1f77b4", "#ff7f0e", "#2ca02c")) +
  geom_abline(intercept = 0, slope = 1, color = "gray", linetype = "dashed") +
  labs(title = "ROC Curves Comparison",
       x = "Specificity",
       y = "Sensitivity",
       color = "Model") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom")

# Threshold Tuning
best_coords <- coords(roc_main, "best", best.method = "youden", 
                      ret = c("threshold", "specificity", "sensitivity"))

# Extract the numeric threshold safely (taking the first if there are ties)
optimal_threshold <- as.numeric(best_coords$threshold[1])
print(paste("Optimal Threshold found:", round(optimal_threshold, 4)))

# Generate predictions using this new threshold
preds_tuned <- factor(ifelse(probs_main > optimal_threshold, "Diabetes", "No_Diabetes"),
                      levels = c("No_Diabetes", "Diabetes"))
```


```{r}
# Calculate confusion matrix stats
cm_tuned <- confusionMatrix(preds_tuned, diabetes_test$diabetes_binary, positive = "Diabetes")
print(cm_tuned)

# Visualizing the Confusion Matrix as a Heatmap
cm_data <- as.data.frame(cm_tuned$table)

ggplot(cm_data, aes(x = Prediction, y = Reference, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), size = 6, color = "white") +
  scale_fill_gradient(low = "#8ebad9", high = "#0b406c") +
  labs(title = paste("Confusion Matrix (Threshold =", round(optimal_threshold, 3), ")"),
       subtitle = "Optimized for Screening (Higher Sensitivity)",
       x = "Predicted Label",
       y = "Actual Label") +
  theme_minimal(base_size = 14)
```


```{r}
# Define the variables we want to keep and how to rename them
# We map the R 'term' to a 'Clean_Name'
plot_data <- tidy(logit_main) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    # Create a nice label for each technical term
    Clean_Name = case_when(
      # Medical Variables
      str_detect(term, "high_bp") ~ "High Blood Pressure",
      str_detect(term, "high_chol") ~ "High Cholesterol",
      str_detect(term, "bmi_category\\.L") ~ "BMI Category (Increasing)", 
      str_detect(term, "bmi") & !str_detect(term, "category") ~ "BMI (Continuous)",
      
      # Behavioral Variables
      str_detect(term, "smoker") ~ "Smoker",
      # Note: Fruits, Veggies, Alcohol were NOT in the logit_main model formula 
      # so they won't appear here unless the model is refit. 
      # We will display what is available from the current model.
      
      # Demographic Variables
      str_detect(term, "age_cat\\.L") ~ "Age (Older)", # Keep only Linear (.L)
      str_detect(term, "income_cat\\.L") ~ "Income (Higher)", # Keep only Linear (.L)
      str_detect(term, "sexMale") ~ "Sex (Male)",
      
      # Everything else (Education, GenHlth, etc.) gets NA so we can drop it
      TRUE ~ NA_character_
    )
  ) %>%
  # Remove variables we didn't define above (GenHlth, Education, .Q, .C terms)
  filter(!is.na(Clean_Name)) %>% 
  mutate(abs_effect = abs(estimate)) %>%
  arrange(desc(abs_effect))

# Plot the filtered top predictors
ggplot(plot_data, aes(x = abs_effect, y = reorder(Clean_Name, abs_effect))) +
  geom_col(fill = "#4C9F70", width = 0.7) +
  geom_text(aes(label = round(abs_effect, 2)), hjust = -0.2, size = 4) + # Add values
  labs(title = "Key Predictors of Diabetes",
       subtitle = "Effect Size (Absolute Log-Odds)",
       x = "Impact on Diabetes Risk",
       y = NULL) +
  theme_minimal(base_size = 14) +
  theme(plot.margin = margin(10, 30, 10, 10)) + # Add room for text labels
  xlim(0, max(plot_data$abs_effect) * 1.1) # Expand x-axis slightly
```


```{r, echo=FALSE}

# Calibration Curve

cal_data <- tibble(
  obs = ifelse(diabetes_test$diabetes_binary == "Diabetes", 1, 0),
  pred = probs_main
) %>%
  mutate(bin = ntile(pred, 10)) %>%
  group_by(bin) %>%
  summarise(mean_pred = mean(pred),
            mean_obs = mean(obs))

ggplot(cal_data, aes(x = mean_pred, y = mean_obs)) +
  geom_point(size = 3, color = "#2C7BB6") +
  geom_line(color = "#2C7BB6") +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray") +
  labs(title = "Calibration Curve (Main Effects)",
       x = "Predicted Probability",
       y = "Observed Proportion") +
  theme_minimal(base_size = 14)
```
