---
title: "Task 1: Data Cleaning and Processing"
author: "cl4775"
date: "December 2025"
output: github_document
---

# P8105 Final Project: Diabetes Health Indicators
 
**Task 1: Data Cleaning, Variable Processing, and Train/Test Dataset Creation**

Dataset: CDC BRFSS 2015 Diabetes Health Indicators  
Source: https://www.kaggle.com/datasets/alexteboul/diabetes-health-indicators-dataset

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Load Required Libraries

```{r load_libraries}
library(tidyverse)
library(caret)
library(skimr)
library(janitor)
library(corrplot)
library(DataExplorer)

# Set seed for reproducibility
set.seed(123)
```

## 1. Data Import

The dataset has three versions:
- `diabetes_012_health_indicators_BRFSS2015.csv` (3 classes: 0, 1, 2)
- `diabetes_binary_5050split_health_indicators_BRFSS2015.csv` (balanced binary)
- `diabetes_binary_health_indicators_BRFSS2015.csv` (full binary)

For this project, we'll use the full binary version for more data.

```{r import_data}
diabetes_raw <- read_csv("diabetes_binary_health_indicators_BRFSS2015.csv")

glimpse(diabetes_raw)
```

## 2. Initial Data Exploration

```{r explore_data}
# Check for missing values
colSums(is.na(diabetes_raw))

# Summary statistics
summary(diabetes_raw)

# Check for duplicates
sum(duplicated(diabetes_raw))

# Examine outcome variable distribution
table(diabetes_raw$Diabetes_binary)
prop.table(table(diabetes_raw$Diabetes_binary))
```

## 3. Data Cleaning

```{r clean_data}
diabetes_clean <- diabetes_raw %>%
  # Clean column names
  clean_names() %>%
  
  # Fix the heart disease variable name
  rename(heart_disease_or_attack = heart_diseaseor_attack) %>%
  
  # Remove duplicates if any
  distinct() %>%
  
  # Remove rows with missing values (if any)
  drop_na()

glimpse(diabetes_clean)
```

## 4. Variable Processing and Feature Engineering

### Variable Descriptions

According to the dataset documentation:

1. **Diabetes_binary**: 0 = no diabetes, 1 = prediabetes or diabetes
2. **HighBP**: 0 = no, 1 = yes
3. **HighChol**: 0 = no, 1 = yes
4. **CholCheck**: 0 = no, 1 = yes (cholesterol check in 5 years)
5. **BMI**: Body Mass Index
6. **Smoker**: 0 = no, 1 = yes
7. **Stroke**: 0 = no, 1 = yes
8. **HeartDiseaseorAttack**: 0 = no, 1 = yes
9. **PhysActivity**: 0 = no, 1 = yes
10. **Fruits**: 0 = no, 1 = yes (consume fruit 1+ times per day)
11. **Veggies**: 0 = no, 1 = yes (consume vegetables 1+ times per day)
12. **HvyAlcoholConsump**: 0 = no, 1 = yes
13. **AnyHealthcare**: 0 = no, 1 = yes
14. **NoDocbcCost**: 0 = no, 1 = yes (couldn't see doctor due to cost)
15. **GenHlth**: 1-5 scale (1 = excellent, 5 = poor)
16. **MentHlth**: days of poor mental health in past 30 days (0-30)
17. **PhysHlth**: days of poor physical health in past 30 days (0-30)
18. **DiffWalk**: 0 = no, 1 = yes (difficulty walking or climbing stairs)
19. **Sex**: 0 = female, 1 = male
20. **Age**: 1-13 (age categories)
21. **Education**: 1-6 (education level)
22. **Income**: 1-8 (income categories)

### Processing

```{r process_variables}
diabetes_processed <- diabetes_clean %>%
  mutate(
    # Convert outcome to factor
    diabetes_binary = factor(diabetes_binary, 
                             levels = c(0, 1),
                             labels = c("No_Diabetes", "Diabetes")),
    
    # Convert binary variables to factors
    high_bp = factor(high_bp, levels = c(0, 1), labels = c("No", "Yes")),
    high_chol = factor(high_chol, levels = c(0, 1), labels = c("No", "Yes")),
    chol_check = factor(chol_check, levels = c(0, 1), labels = c("No", "Yes")),
    smoker = factor(smoker, levels = c(0, 1), labels = c("No", "Yes")),
    stroke = factor(stroke, levels = c(0, 1), labels = c("No", "Yes")),
    heart_disease_or_attack = factor(heart_disease_or_attack, 
                                     levels = c(0, 1), 
                                     labels = c("No", "Yes")),
    phys_activity = factor(phys_activity, levels = c(0, 1), labels = c("No", "Yes")),
    fruits = factor(fruits, levels = c(0, 1), labels = c("No", "Yes")),
    veggies = factor(veggies, levels = c(0, 1), labels = c("No", "Yes")),
    hvy_alcohol_consump = factor(hvy_alcohol_consump, 
                                 levels = c(0, 1), 
                                 labels = c("No", "Yes")),
    any_healthcare = factor(any_healthcare, levels = c(0, 1), labels = c("No", "Yes")),
    no_docbc_cost = factor(no_docbc_cost, levels = c(0, 1), labels = c("No", "Yes")),
    diff_walk = factor(diff_walk, levels = c(0, 1), labels = c("No", "Yes")),
    sex = factor(sex, levels = c(0, 1), labels = c("Female", "Male")),
    
    # Convert ordinal variables to factors
    gen_hlth_cat = factor(gen_hlth, 
                          levels = 1:5,
                          labels = c("Excellent", "Very Good", "Good", "Fair", "Poor"),
                          ordered = TRUE),
    
    # Create age groups
    age_cat = factor(age,
                     levels = 1:13,
                     labels = c("18-24", "25-29", "30-34", "35-39", "40-44",
                                "45-49", "50-54", "55-59", "60-64", "65-69",
                                "70-74", "75-79", "80+"),
                     ordered = TRUE),
    
    # Create education categories
    education_cat = factor(education,
                           levels = 1:6,
                           labels = c("Never/K-8", "Grades 9-11", "Grade 12/GED",
                                      "College 1-3 yrs", "College 4+ yrs", "Unknown"),
                           ordered = TRUE),
    
    # Create income categories
    income_cat = factor(income,
                        levels = 1:8,
                        labels = c("<$10k", "$10k-$15k", "$15k-$20k", "$20k-$25k",
                                   "$25k-$35k", "$35k-$50k", "$50k-$75k", "$75k+"),
                        ordered = TRUE),
    
    # Create BMI categories
    bmi_category = case_when(
      bmi < 18.5 ~ "Underweight",
      bmi >= 18.5 & bmi < 25 ~ "Normal",
      bmi >= 25 & bmi < 30 ~ "Overweight",
      bmi >= 30 ~ "Obese"
    ) %>% factor(levels = c("Underweight", "Normal", "Overweight", "Obese"),
                 ordered = TRUE),
    
    # Create mental health categories
    ment_hlth_cat = case_when(
      ment_hlth == 0 ~ "None",
      ment_hlth <= 7 ~ "1-7 days",
      ment_hlth <= 14 ~ "8-14 days",
      ment_hlth <= 30 ~ "15-30 days"
    ) %>% factor(levels = c("None", "1-7 days", "8-14 days", "15-30 days"),
                 ordered = TRUE),
    
    # Create physical health categories
    phys_hlth_cat = case_when(
      phys_hlth == 0 ~ "None",
      phys_hlth <= 7 ~ "1-7 days",
      phys_hlth <= 14 ~ "8-14 days",
      phys_hlth <= 30 ~ "15-30 days"
    ) %>% factor(levels = c("None", "1-7 days", "8-14 days", "15-30 days"),
                 ordered = TRUE),
    
    # Create composite health score
    health_risk_score = as.numeric(high_bp == "Yes") +
      as.numeric(high_chol == "Yes") +
      as.numeric(stroke == "Yes") +
      as.numeric(heart_disease_or_attack == "Yes") +
      as.numeric(smoker == "Yes") +
      as.numeric(hvy_alcohol_consump == "Yes") +
      (gen_hlth - 1) / 4,
    
    # Create healthy lifestyle score
    lifestyle_score = as.numeric(phys_activity == "Yes") +
      as.numeric(fruits == "Yes") +
      as.numeric(veggies == "Yes")
  )

glimpse(diabetes_processed)
```

## 5. Data Quality Checks

```{r quality_checks}
# Check for any issues in processed data
sum(is.na(diabetes_processed))

# Verify factor levels
table(diabetes_processed$diabetes_binary)

# Check continuous variables for outliers
diabetes_processed %>%
  select(bmi, ment_hlth, phys_hlth, health_risk_score, lifestyle_score) %>%
  summary()
```

## 6. Train/Test Split

Create stratified train/test split (80/20) using stratified sampling to maintain outcome distribution.

```{r train_test_split}
# Set seed again for reproducibility
set.seed(123)

# Create partition index
train_index <- createDataPartition(
  diabetes_processed$diabetes_binary,
  p = 0.80,
  list = FALSE,
  times = 1
)

# Create training and testing datasets
diabetes_train <- diabetes_processed[train_index, ]
diabetes_test <- diabetes_processed[-train_index, ]

# Verify split proportions
nrow(diabetes_train)
nrow(diabetes_test)

# Check outcome distribution in train and test sets
prop.table(table(diabetes_train$diabetes_binary))
prop.table(table(diabetes_test$diabetes_binary))
```

## 7. Export Cleaned Datasets

```{r export_data}
# Create output directory if it doesn't exist
if (!dir.exists("data")) {
  dir.create("data")
}

# Save processed datasets
write_csv(diabetes_processed, "data/diabetes_processed.csv")
write_csv(diabetes_train, "data/diabetes_train.csv")
write_csv(diabetes_test, "data/diabetes_test.csv")

# Also save as RDS for faster loading in R
saveRDS(diabetes_processed, "data/diabetes_processed.rds")
saveRDS(diabetes_train, "data/diabetes_train.rds")
saveRDS(diabetes_test, "data/diabetes_test.rds")
```

## 8. Generate Data Dictionary

```{r data_dictionary}
# Create data dictionary
data_dictionary <- tribble(
  ~Variable, ~Type, ~Description, ~Values,
  "diabetes_binary", "Factor", "Diabetes diagnosis", "No_Diabetes, Diabetes",
  "high_bp", "Factor", "High blood pressure", "No, Yes",
  "high_chol", "Factor", "High cholesterol", "No, Yes",
  "chol_check", "Factor", "Cholesterol check in past 5 years", "No, Yes",
  "bmi", "Numeric", "Body Mass Index", "Continuous",
  "smoker", "Factor", "Smoked at least 100 cigarettes in life", "No, Yes",
  "stroke", "Factor", "History of stroke", "No, Yes",
  "heart_disease_or_attack", "Factor", "Coronary heart disease or MI", "No, Yes",
  "phys_activity", "Factor", "Physical activity in past 30 days", "No, Yes",
  "fruits", "Factor", "Consume fruit 1+ times per day", "No, Yes",
  "veggies", "Factor", "Consume vegetables 1+ times per day", "No, Yes",
  "hvy_alcohol_consump", "Factor", "Heavy alcohol consumption", "No, Yes",
  "any_healthcare", "Factor", "Have any healthcare coverage", "No, Yes",
  "no_docbc_cost", "Factor", "Could not see doctor due to cost", "No, Yes",
  "gen_hlth", "Numeric", "General health rating", "1-5 (1=excellent, 5=poor)",
  "ment_hlth", "Numeric", "Days of poor mental health (past 30)", "0-30",
  "phys_hlth", "Numeric", "Days of poor physical health (past 30)", "0-30",
  "diff_walk", "Factor", "Difficulty walking or climbing stairs", "No, Yes",
  "sex", "Factor", "Biological sex", "Female, Male",
  "age", "Numeric", "Age category", "1-13 (see age_cat for labels)",
  "education", "Numeric", "Education level", "1-6",
  "income", "Numeric", "Income level", "1-8",
  "bmi_category", "Factor", "BMI category", "Underweight, Normal, Overweight, Obese",
  "health_risk_score", "Numeric", "Composite health risk score", "Continuous (0-7)",
  "lifestyle_score", "Numeric", "Healthy lifestyle score", "0-3"
)

# Save data dictionary
write_csv(data_dictionary, "data/data_dictionary.csv")

# Display dictionary
data_dictionary
```


## Exploratory analysis


```{r}
# Load processed data
diabetes <- read_csv("data/diabetes_processed.csv") %>%
  mutate(
    diabetes_binary = factor(diabetes_binary,
                             levels = c("No_Diabetes", "Diabetes"))
  )
```

# Research Questions

In this project we would like to focus on these three questions: 

1. Which demographic, lifestyle, and clinical characteristics are most strongly associated with diabetes?

2. How effectively can logistic regression models predict diabetes status using only self-reported survey information?

3. Does model complexity like interaction terms improve predictive performance?

# Data

We use the cleaned [2015 BRFSS dataset](https://www.kaggle.com/datasets/alexteboul/diabetes-health-indicators-dataset/data) (253,680 adults, 22 variables), focusing on variables related to demographics, lifestyle, and self-reported health.

- **Population:** Adults participating in the 2015 BRFSS survey.
- **Outcome:** Self-reported diabetes status (binary outcome used for prediction).
- **Main predictors:** BMI, age category, general health rating, blood pressure, cholesterol, smoking status, physical activity, sex, and socioeconomic indicators

## Key preprocessing steps:

- No missing data were present (already cleaned by BRFSS).

- Created interpretable new variables according to [dataset notebook](https://www.kaggle.com/code/alexteboul/diabetes-health-indicators-dataset-notebook) and [codebook](https://www.cdc.gov/brfss/annual_data/2015/pdf/codebook15_llcp.pdf). 

  - BMI categories

  - Mental & physical health categories

  - Composite lifestyle and health-risk scores

  - Converted categorical variables into meaningful factor labels

- Train/test Split

  - 80/20 stratified split is used to preserve diabetes prevalence.
  
  - Training set: approximately 202,944 observations (for EDA and model fitting)
  
  - Test set: approximately 50,736 observations (for final evaluation only)

- Fixed random seed (123) for reproducibility

## Exploratory analysis revealed:

- **Strong class imbalance (far more non-diabetic respondents).**

```{r}
# Distribution of Diabetes Status
diabetes %>%
  ggplot(aes(x = diabetes_binary, fill = diabetes_binary)) +
  geom_bar(width = 0.55, alpha = 0.9, color = "white") +
  scale_fill_manual(values = c("#5DA5DA", "#F17CB0")) + 
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(
    title = "Distribution of Diabetes Status",
    x = "Diabetes Status",
    y = "Count"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
    axis.title.x = element_text(size = 14, margin = margin(t = 10)),
    axis.title.y = element_text(size = 14, margin = margin(r = 10)),
    axis.text = element_text(size = 12),
    legend.position = "none",
    plot.margin = margin(15, 15, 15, 15)
  )

```

According to this plot, diabetes individual included is much less than non-diabetes. This may lead to strong class imbalance. 

```{r}
# BMI Histogram
diabetes %>%
  ggplot(aes(x = bmi)) +
  geom_histogram(bins = 40, fill = "#6495ED", alpha = 0.8) +
  labs(
    title = "Distribution of BMI",
    x = "BMI",
    y = "Count"
  ) +
  theme_minimal()
```

The BMI in this sample is right-skewed, with most values falling between 20 and 40. 

The concentration of observations in the overweight and obese range reflects the population distribution in this dataset. A smaller number of individuals have very high BMI values, forming a long right tail.

```{r}
# Age Category Distribution
diabetes %>%
  ggplot(aes(x = age_cat)) +
  geom_bar(fill = "#8B5FBF", alpha = 0.8) +
  labs(
    title = "Age Category Distribution",
    x = "Age Category",
    y = "Count"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Left-skewness in this plot shows that the dataset contains far more middle-aged and older adults than younger respondents. Counts increase steadily from the 18–24 group and peak between ages 55–69, before gradually declining in the 70+ categories. This uneven age distribution is important because diabetes risk rises with age. This means that the sample naturally contains more individuals from higher-risk age groups.

- **Diabetic individuals tend to have higher BMI, poor general health, and older age.**

```{r}
# BMI by Diabetes Status
diabetes %>%
  ggplot(aes(x = diabetes_binary, y = bmi, fill = diabetes_binary)) +
  geom_boxplot(alpha = 0.85) +
  scale_fill_manual(values = c("lightblue", "lightpink")) +
  labs(
    title = "BMI by Diabetes Status",
    x = "Diabetes Status",
    y = "BMI"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

The plot shows that individuals included with diabetes tend to have higher BMI values than those without diabetes. Both the median and overall distribution of diabetes group shifted upward. 

```{r}
# Age Category (Numeric) by Diabetes Status
diabetes %>%
  ggplot(aes(x = diabetes_binary, y = age, fill = diabetes_binary)) +
  geom_boxplot(alpha = 0.85) +
  labs(
    title = "Age Category by Diabetes Status",
    x = "Diabetes Status",
    y = "Age Category (1 = 18–24 … 13 = 80+)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

The plot shows that the median age category is higher for the diabetes group, and their overall distribution is shifted toward older age ranges.

```{r}
# General Health Score by Diabetes Status
diabetes %>%
  ggplot(aes(x = diabetes_binary, y = gen_hlth, fill = diabetes_binary)) +
  geom_boxplot(alpha = 0.85) +
  scale_fill_manual(values = c("#4C9F70", "#C94C4C")) +
  labs(
    title = "General Health Score by Diabetes Status",
    x = "Diabetes Status",
    y = "General Health (1 = Excellent, 5 = Poor)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

The plot shows that individuals with diabetes generally report worse overall health than those without diabetes. The diabetes group has a higher median health score, which indicates poorer health and a distribution shifted toward less favorable health ratings.

- **Income distribution is skewed toward higher brackets, which may influence access to care.**

```{r}
# Income Category Distribution
diabetes %>%
  ggplot(aes(x = income_cat)) +
  geom_bar(fill = "#FFA500", alpha = 0.8) +
  labs(
    title = "Income Level Distribution",
    x = "Income Category",
    y = "Count"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
# Correlation heat map
# Select numeric variables
numeric_vars <- diabetes %>%
  select(bmi, ment_hlth, phys_hlth, gen_hlth, income, age)

# Compute and tidy correlation matrix
cor_df <- numeric_vars %>%
  cor(use = "complete.obs") %>%
  as.data.frame() %>%
  rownames_to_column("var1") %>%
  pivot_longer(-var1, names_to = "var2", values_to = "cor")

# Heatmap with correlation values
cor_df %>%
  ggplot(aes(var1, var2, fill = cor)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(
    low = "#4575b4",
    high = "#d73027",
    mid = "white",
    midpoint = 0,
    limits = c(-1, 1)
  ) +
  geom_text(aes(label = round(cor, 2)), size = 3) +
  labs(
    title = "Correlation Heatmap of Numeric Predictors",
    x = "",
    y = "",
    fill = "Correlation"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The heatmap shows that correlations among the numeric predictors are generally weak to moderate, indicating low multicollinearity and supporting their joint use in logistic regression models. 

Variables related to overall health, such as general health and physical or mental health days, are moderately correlated as expected.

While BMI and age show only weak relationships with other predictors. 
Overall, the plot confirms that each variable contributes distinct information for diabetes risk prediction.


 
